---
title: "Toxicity TApp Group D"
author: "Group D"
date: "2024-045-22"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

- **Davide Albiero** davide.albiero@studenti.unipd.it
- **Nazanin Ghorbani** nazanin.ghorbani@studenti.unipd.it
- **Raman Yudzeshka** raman.yudzeshka@studenti.unipd.it
- **Niccolò Zenaro** niccolo.zenaro@studenti.unipd.it
- **Luca Marchiori** luca.marchiori.3@studenti.unipd.it
- **Mohammad Khosravi** mohammad.khosravi.1@studenti.unipd.it

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r include=FALSE}
# Load libraries
library(caret)
library(ggplot2)
library(corrplot)
library(mgcv)
library(tidyr)
library(dplyr)
library(ISLR2)
library(GGally)
#Clear the workspace
rm(list=ls())
```


## Data import
```{r}
# Load the training data
train <- read.csv("toxicity_train.csv", stringsAsFactors=F)
# Load the test data
test <- read.csv("toxicity_validation.csv", stringsAsFactors=F)
set.seed(123)
```

## Data Visualization
```{r}
summary(train)

ggpairs(train,
        upper = list(continuous = wrap("cor", size = 2.5, color = "blue")),
  lower = list(continuous = wrap("points", size = 0.8, alpha = 0.7, color = "red")),
  diag = list(continuous = wrap("densityDiag", fill = "lightblue", alpha = 0.7)),
  axisLabels = "show")
```
By looking the diagonal, it appears that the feature `LC50`, `MLOGP` and `RDCHI` follow a normal distribution, while the others features has a tail.
```{r}
#TODO what's going on
ggplot(train, aes(x =nN, y=LC50)) + geom_point() + theme_minimal()
ggplot(train, aes(x =H050, y=LC50)) + geom_point() + theme_minimal()
ggplot(train, aes(x =C040, y=LC50)) + geom_point() + theme_minimal()

# Seams like there is a linear relationship between LC50 and MLOGP
ggplot(train, aes(x =MLOGP, y=LC50)) + geom_point() + theme_minimal()
```

### Task 3

For this task, we train the model multiple times by increasing at each train the polynomial degree (from 1 to 5). The RMSE of each model is calculated after the cross-validation. The cross validation is used to ensure that the calculated RMSE is a good quality and not biased measure.

The results show that the 2-degree polynomial is the one with the lowest RMSE. A plot that compares the RMSE and the degree of polynomial is shown. In the end we also plot the final model function on the data scatterplot.
```{r}
# Train a model by trying multiple polynomial degrees. Save the models and their cross-validated RMSEs in a list and choose the best model based on the RMSE.

degrees <- 1:7
models <- list()
rmses <- c()

fitControl <- trainControl(method = "cv", number = 20)

for (degree in degrees){
  # Build up the formula with the selected degree
  formula <- as.formula(paste("LC50 ~ poly(RDCHI, ", degree, ")", sep=""))
  
  # Train the model
  fit <- caret::train(formula, data=train, method="lm", trControl=fitControl, preProcess=c("center", "scale"))
  
  # Get the cross-validated RMSE
  fitRmse <- fit$results$RMSE
  # Save the model RMSE
  rmses <- c(rmses, fitRmse)
}

# Plot the flexibility of the model vs RMSE
ggplot(data.frame(degree=degrees, RMSE=rmses), aes(x=degree, y=RMSE)) +
  geom_point() +
  geom_line() +
  xlab("Degree of polynomial") + ylab("RMSE") +
  theme_minimal()

# Choose the best model
bestDegree <- degrees[which.min(rmses)]
bestDegree

# Train the best model again (just for the sake of plotting it)
bestFormula <- as.formula(paste("LC50 ~ poly(RDCHI, ", bestDegree, ")", sep=""))
bestFit <- lm(bestFormula, data=train, method="lm", trControl=fitControl)

# Plot the data and the best model
ggplot(train, aes(x=RDCHI, y=LC50)) +
  geom_point(size=0.5, color="#34495e") +
  geom_line(aes(y=predict(bestFit), color="#c0392b"), show.legend = FALSE) +
  theme_minimal()
```

### Task 4
In this task, we aimed to compare the performance of a step function model with 29 knots and a cubic spline model with 26 knots, both resulting in 30 DOF.
```{r}
# let's use 30 DOF and compare step function and cubic spline
# then it'll be 30 knots for the step function and 26 for the cubic spline

knots <- quantile(train$RDCHI, probs = seq(0, 1, length.out = 28))

# Step function
train$RDCHI_step <- cut(train$RDCHI, breaks=30, labels=FALSE)

# Define the model formulas
step_model_formula <- LC50 ~ factor(RDCHI_step)

str_knots <- paste("c(", noquote(paste(knots[-c(1, length(knots))], collapse = ", ")), ")", sep = "")
spline_model_formula <- as.formula(paste("LC50 ~ splines::bs(RDCHI, knots = ", str_knots, ", degree = 3)", sep = ""))
#spline_model_formula

# Define train control with 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 10)

# Fit the models and calculate RMSE using cross-validation
model_step <- train(step_model_formula, data = train, method = "lm", trControl = ctrl)
model_spline <- train(spline_model_formula, data = train, method = "lm", trControl = ctrl)

#summary(model_spline)


# Predicting
train$predicted_LC50_spline <- predict(model_spline, newdata=train)
train$predicted_LC50_step <- predict(model_step, newdata=train)

# Print RMSE for both models
cat("RMSE for Step Function Model (CV): ", model_step$results$RMSE, "\n")
cat("RMSE for Cubic Spline Model (CV): ", model_spline$results$RMSE, "\n")
```
The step function model slightly outperformed the cubic spline model (RMSE: 1.441 vs. 1.510) due to its ability to better capture abrupt changes in the data.
```{r}
ggplot() +
  # Train data
  geom_point(data=train, aes(x = RDCHI, y = LC50), alpha = 0.5) +
  # Step function predictions
  stat_summary_bin(data=train, aes(x = RDCHI, y = predicted_LC50_step), 
                   bins = 30, geom = "step", color = "red", linewidth = 1, alpha = 0.8) +
  # Cubic spline predictions
  geom_line(data=train, aes(x = RDCHI, y = predicted_LC50_spline), 
            color = "blue", linewidth = 1, alpha = 0.8) +
  labs(title = "Step Function vs Cubic Spline Regression", 
       x = "RDCHI", y = "LC50") +
  theme_minimal()
```

### Task 5
First, we need to find the best degrees of freedom for the model. We use the cross-validation score to find the best degrees of freedom. In particular we use PRESS (PREdiction Sum of Squares) as the cross-validation score.
```{r}
max_df = 30
res <- list()
for (i in 2:max_df) {
  spline <- smooth.spline(train$RDCHI, train$LC50, df = i)
  # print(paste("df:", i, "Cross-validation score:", spline$cv.crit))
  res <- c(res, spline$cv.crit)
}
best_df <- which.min(res)
```
We plot the degrees of freedom on the x-axis and the cross-validation score on the y-axis. The best model is chosen based on the minimum cross-validation score.
```{r echo=FALSE}
plot(2:max_df, res, type = "l", xlab = "Degrees of freedom", ylab = "Cross-validation score")
points(best_df, res[best_df], pch = 19, col = "red")
legend("topright", legend = c(paste("Best df:", best_df, sep = "")), col = c("red"), pch = 19)
```
We plot the data, the basis linear model and the chosen spline model with the best degrees of freedom.
The model is chosen based on the minimum cross-validation score.
```{r}
reg <- lm(LC50 ~ RDCHI, data=train)
plot(train$RDCHI, train$LC50, xlab="RDCHI", ylab="LC50", pch=20, col="black")
abline(reg, col="blue")
#fit a spline with x degrees of freedom
spline <- smooth.spline(train$RDCHI, train$LC50, df=which.min(res))
lines(spline, col="red")
legend("topright", legend=c("Linear", "Spline"), col=c("blue", "red"), lty=1)
```

### Task 6
Natural cubic spline:
GAM is fitted using natural cubic splines to predict LC50 based on various chemical descriptors. The model is tuned to handle predictors with fewer unique values by adjusting the number of knots. Partial effects of each predictor are extracted and visualized using scatter plots, providing insights into their individual contributions to the LC50 prediction. By adjusting the number of knots, the model captures the complexity of each predictor's contribution to the LC50 prediction, providing a detailed understanding of the data.
```{r}
# Check the number of unique values in each predictor
sapply(train, function(x) length(unique(x)))

# Fit a GAM model using natural cubic splines with reduced knots for each predictor
gam_model <- gam(LC50 ~ 
                   s(TPSA, bs = "cs", k = 20) +   # Less than 193
                   s(SAacc, bs = "cs", k = 20) +  # Less than 182
                   s(H050, bs = "cs", k = 9) +    # Less than 10
                   s(MLOGP, bs = "cs", k = 20) +  # Less than 349 (k can be higher but 20 is reasonable)
                   s(RDCHI, bs = "cs", k = 20) +  # Less than 297 (k can be higher but 20 is reasonable)
                   s(GATS1p, bs = "cs", k = 20) + # Less than 349 (k can be higher but 20 is reasonable)
                   s(nN, bs = "cs", k = 9) +      # Less than 9
                   s(C040, bs = "cs", k = 6),     # Less than 6
                 data = train)

# Summarize the model
summary(gam_model)

# Extract partial effects of each predictor
partial_effects <- predict(gam_model, type = "terms", se.fit = TRUE)
partial_effects_df <- as.data.frame(partial_effects$fit)
colnames(partial_effects_df) <- gsub("\\.s\\(.*\\)", "", colnames(partial_effects_df))
partial_effects_df$LC50 <- train$LC50

# Convert the data to long format for ggplot
partial_effects_long <- partial_effects_df %>%
  pivot_longer(cols = -LC50, names_to = "Predictor", values_to = "Effect")

# Plot the effects of each predictor using ggplot2
ggplot(partial_effects_long, aes(x = Effect, y = LC50)) +
  geom_point() +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(title = "Partial Effects of Predictors on LC50 using natural cubic splines",
       x = "Partial Effect",
       y = "LC50") +
  theme_minimal()

```

Step Function:
A Generalized Additive Model (GAM) was fitted to predict LC50 using step functions
The predictors are binned, and their partial effects are extracted and visualized using scatter plots. This approach allows for a flexible, non-linear relationship between the predictors and LC50, providing a more accurate representation of the data.

```{r}
# Check the number of unique values in each predictor
unique_counts <- sapply(train, function(x) length(unique(x)))
print(unique_counts)

# Define the number of bins for each predictor
bins <- list(TPSA = 10, SAacc = 10, H050 = 4, MLOGP = 10, RDCHI = 10, GATS1p = 10, nN = 4, C040 = 3)

# Create binned predictors in the training set
train_binned <- train %>%
  mutate(across(names(bins), ~ cut(., breaks = bins[[cur_column()]], labels = FALSE)))

# Fit a GAM model using step functions (categorical predictors)
gam_model_step <- gam(LC50 ~ 
                        factor(TPSA) + 
                        factor(SAacc) + 
                        factor(H050) + 
                        factor(MLOGP) + 
                        factor(RDCHI) + 
                        factor(GATS1p) + 
                        factor(nN) + 
                        factor(C040),
                      data = train_binned)

# Summarize the model
summary(gam_model_step)

# Extract partial effects of each predictor
partial_effects <- predict(gam_model_step, type = "terms", se.fit = TRUE)

# Manually assign column names to partial effects dataframe
partial_effects_df <- as.data.frame(partial_effects$fit)
colnames(partial_effects_df) <- c("TPSA", "SAacc", "H050", "MLOGP", "RDCHI", "GATS1p", "nN", "C040")
partial_effects_df$LC50 <- train$LC50

# Print column names to ensure they are correct
print(colnames(partial_effects_df))

# Convert the data to long format for ggplot
partial_effects_long <- partial_effects_df %>%
  pivot_longer(cols = -LC50, names_to = "Predictor", values_to = "Effect")

# Plot the effects of each predictor using ggplot2
ggplot(partial_effects_long, aes(x = Effect, y = LC50)) +
  geom_point() +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(title = "Partial Effects of Predictors on LC50 using Step Functions",
       x = "Partial Effect",
       y = "LC50") +
  theme_minimal()
```
Smoothing splines:

```{r}
# Check the number of unique values in each predictor
sapply(train, function(x) length(unique(x)))

# Fit a GAM model using smoothing splines with reduced knots for each predictor
gam_model <- gam(LC50 ~ 
                   s(TPSA, bs = "cr", k = 20) +   # Less than 193
                   s(SAacc, bs = "cr", k = 20) +  # Less than 182
                   s(H050, bs = "cr", k = 9) +    # Less than 10
                   s(MLOGP, bs = "cr", k = 20) +  # Less than 349 (k can be higher but 20 is reasonable)
                   s(RDCHI, bs = "cr", k = 20) +  # Less than 297 (k can be higher but 20 is reasonable)
                   s(GATS1p, bs = "cr", k = 20) + # Less than 349 (k can be higher but 20 is reasonable)
                   s(nN, bs = "cr", k = 9) +      # Less than 9
                   s(C040, bs = "cr", k = 6),     # Less than 6
                 data = train)

# Summarize the model
summary(gam_model)

# Extract partial effects of each predictor
partial_effects <- predict(gam_model, type = "terms", se.fit = TRUE)
partial_effects_df <- as.data.frame(partial_effects$fit)
colnames(partial_effects_df) <- gsub("\\.s\\(.*\\)", "", colnames(partial_effects_df))
partial_effects_df$LC50 <- train$LC50

# Convert the data to long format for ggplot
partial_effects_long <- partial_effects_df %>%
  pivot_longer(cols = -LC50, names_to = "Predictor", values_to = "Effect")

# Plot the effects of each predictor using ggplot2
ggplot(partial_effects_long, aes(x = Effect, y = LC50)) +
  geom_point() +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(title = "Partial Effects of Predictors on LC50 using smoothing splines",
       x = "Partial Effect",
       y = "LC50") +
  theme_minimal()
```

### Regression Spline (similar to smoothing spline)
```{r}
# Check the number of unique values in each predictor
sapply(train, function(x) length(unique(x)))

# Fit a GAM model using Regression spline
gam_model <- gam(LC50 ~ 
                 s(TPSA, bs = "cr", k = 20) +
                 s(SAacc, bs = "cr", k = 20) +
                 s(H050, bs = "cr", k = 9) + 
                 s(MLOGP, bs = "cr", k = 20) + 
                 s(RDCHI, bs = "cr", k = 20) + 
                 s(GATS1p, bs = "cr", k = 20) + 
                 s(nN, bs = "cr", k = 9) + 
                 s(C040, bs = "cr", k = 6),
               data = train, method = "REML") #REML(Restricted Maximum Likelihood) to estimate the optimal amount of                                                    smoothing for each term in the model and avoid over fitting


# Summary of the GAM model
summary(gam_model)

# Extract partial effects of each predictor
partial_effects <- predict(gam_model, type = "terms", se.fit = TRUE)
partial_effects_df <- as.data.frame(partial_effects$fit)
colnames(partial_effects_df) <- gsub("\\.s\\(.*\\)", "", colnames(partial_effects_df))
partial_effects_df$LC50 <- train$LC50

# Convert the data to long format for ggplot
partial_effects_long <- partial_effects_df %>%
  pivot_longer(cols = -LC50, names_to = "Predictor", values_to = "Effect")

# Plot the effects of each predictor using ggplot2
ggplot(partial_effects_long, aes(x = Effect, y = LC50)) +
  geom_point() +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(title = "Partial Effects of Predictors on LC50 using Regression Splines",
       x = "Partial Effect",
       y = "LC50") +
  theme_minimal()
```


### Polynomial
```{r}
# Fit a GAM model using polynomial terms
n = 3
gam_model <- gam(LC50 ~ 
                 poly(TPSA, n) +
                 poly(SAacc, n) + 
                 poly(H050, n) + 
                 poly(MLOGP, n) + 
                 poly(RDCHI, n) + 
                 poly(GATS1p, n) + 
                 poly(nN, n) + 
                 poly(C040, n),
               data = train, method = 'REML')#REML(Restricted Maximum Likelihood) to estimate the optimal amount of                                                    smoothing for each term in the model and avoid over fitting

# Summary of the GAM model
summary(gam_model)

# Extract partial effects of each predictor
partial_effects <- predict(gam_model, type = "terms", se.fit = TRUE)
partial_effects_df <- as.data.frame(partial_effects$fit)
colnames(partial_effects_df) <- gsub("\\.s\\(.*\\)", "", colnames(partial_effects_df))
partial_effects_df$LC50 <- train$LC50

# Convert the data to long format for ggplot
partial_effects_long <- partial_effects_df %>%
  pivot_longer(cols = -LC50, names_to = "Predictor", values_to = "Effect")

# Plot the effects of each predictor using ggplot2
ggplot(partial_effects_long, aes(x = Effect, y = LC50)) +
  geom_point() +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(title = "Partial Effects of Predictors on LC50 using polynomial",
       x = "Partial Effect",
       y = "LC50") +
  theme_minimal()
```
